Supermartingales and Termination Analysis of Statistical PCF

### 2021-10-01 LICS21/related.tex
" \ref" -> "~\ref"

\cite{DBLP:conf/lics/KobayashiLG19} shows that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
Note that $n$-PHORS is strictly less expressive then order-$n$ call-by-name PPCF:
% (because the underlying recursion schemes are not Turing complete):
for example, the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
%\refExample{complexExample} 
are not definable as PHORS.

The main result of \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our \cref{thm:antitone rankable implies termination}, but in an imperative language. While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion. 
The difference in the exact progress condition is insignificant. A ranking function satisfying either progress condition can easily be converted to satisfy the other.
\akr{Their progress condition, that there are antitone functions $p$ and $d$ such that the ranking function decreases from $x$ by at least $d(x)$ with probability at least $p(x)$, would not have worked directly because it does not cope as easily with the fact that the ranking function may change in value at the intermediate non-$\tY$ reduction steps (provided it does not increase in expectation). However, it is easy enough to convert a ranking function that satisfies either progress condition to the other.}
\lo{Can you replace the highlighted by a brief comment on their discussion of completeness?}
\akr{I would, but there's no space. Their completeness results seem pretty weak anyway.}

@@

\paragraph*{Further directions} We have not yet been able to prove \Cref{conj:antitone} which would imply that \Cref{thm:antitone rankable implies termination} is almost complete.
\begin{conjecture}[Completeness]
\label{conj:antitone}
If every term reachable from a certain PPCF term is AST, then that term is antitone rankable.
\end{conjecture}

The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose. For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples. Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine. A more complete characterisation of which redexes can be reduced without breaking confluence could be interesting.

Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and \Cref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.

>>>>

\cite{DBLP:conf/lics/KobayashiLG19} shows that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
Note that $n$-PHORS is strictly less expressive then order-$n$ call-by-name PPCF:
% (because the underlying recursion schemes are not Turing complete):
for example, the PPCF terms in Examples~\ref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
%\refExample{complexExample} 
are not definable as PHORS.

The main result of \cite{DBLP:journals/pacmpl/McIverMKK18} is similar to our Theorem~\ref{thm:antitone rankable implies termination}, but in an imperative language. While we require that the antitone ranking function decreases for every $\tY$-reduction step, they require that it decreases for every iteration of a certain while loop (the ranking function in that case is defined in the context of a particular loop), which is similar to $\tY$ but limited to tail recursion. 
The difference in the exact progress condition is insignificant. A ranking function satisfying either progress condition can easily be converted to satisfy the other.
\akr{Their progress condition, that there are antitone functions $p$ and $d$ such that the ranking function decreases from $x$ by at least $d(x)$ with probability at least $p(x)$, would not have worked directly because it does not cope as easily with the fact that the ranking function may change in value at the intermediate non-$\tY$ reduction steps (provided it does not increase in expectation). However, it is easy enough to convert a ranking function that satisfies either progress condition to the other.}
\lo{Can you replace the highlighted by a brief comment on their discussion of completeness?}
\akr{I would, but there's no space. Their completeness results seem pretty weak anyway.}

@@

\paragraph*{Further directions} We have not yet been able to prove Conjecture~\ref{conj:antitone} which would imply that Theorem~\ref{thm:antitone rankable implies termination} is almost complete.
\begin{conjecture}[Completeness]
\label{conj:antitone}
If every term reachable from a certain PPCF term is AST, then that term is antitone rankable.
\end{conjecture}

Definition~\ref{def:more general red}, defining redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose. For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples. Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine. A more complete characterisation of which redexes can be reduced without breaking confluence could be interesting.

Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems~\ref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.

### 2021-10-01 LICS21/supermartingales-proofs.tex
Replace "theorem" environment with "therm", to get numbering back.
" \ref" -> "~\ref"

\rankableAndStrictRankable*
We shall obtain the theorem as a corollary of a technical lemma (Lemma \ref{lem:key rankable}).

@@

As an immediate corollary of Lemma \ref{lem:key rankable}, each $f(M_n)$ is integrable.
This concludes the proof of Theorem \ref{thm:rankable and strict rankable}.

@@

\item Further there is a finite bound (say $l$) on the length of such $\transform{M}$-reduction sequences; and $l$ bounds the stopping time $T_1$.
\end{enumerate}
This concludes the proof of Lemma \ref{lem:TnBounded}.

>>>>

\rankableAndStrictRankable*
We shall obtain the theorem as a corollary of a technical lemma (Lemma~\ref{lem:key rankable}).

@@

As an immediate corollary of Lemma~\ref{lem:key rankable}, each $f(M_n)$ is integrable.
This concludes the proof of Theorem~\ref{thm:rankable and strict rankable}.

@@

\item Further there is a finite bound (say $l$) on the length of such $\transform{M}$-reduction sequences; and $l$ bounds the stopping time $T_1$.
\end{enumerate}
This concludes the proof of Lemma~\ref{lem:TnBounded}.

### 2021-10-01 LICS21/supermartingales.tex
Replace "theorem" environment with "therm", to get numbering back.
" \ref" -> "~\ref"

\begin{definition}\rm
\label{def:ranking function}
A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.} and
\begin{enumerate}
    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z)$ where $z$ is not free in $N$
    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$ with $R \to R'$
\end{enumerate}
\iffalse
\lo{@Andrew: As defined, ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value. But for AST analysis, we always require $f$ to satisfy this property. It seems cleaner to include this in the definition.} 
\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}
\lo{OK. Let's leave \Cref{def:ranking function} as it is.}
\fi
We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.

@@

\begin{remark}
Lem.~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
\end{remark}

@@

%\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (\Cref{lem:key rankable}).

@@

\begin{proof}
Let $f$ be a ranking function on $M$.
For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$.
Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
%For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
%It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
Statement (ii) follows at once from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
\end{proof}

Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of PPCF programs.
It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (\Cref{thm:minimal}).

>>>>

\begin{definition}\rm
\label{def:ranking function}
A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem~\ref{thm:minimal} to work neatly.} and
\begin{enumerate}
    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z)$ where $z$ is not free in $N$
    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$ with $R \to R'$
\end{enumerate}
\iffalse
\lo{@Andrew: As defined, ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value. But for AST analysis, we always require $f$ to satisfy this property. It seems cleaner to include this in the definition.} 
\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get Theorem~\ref{thm:minimal} to work neatly.}
\lo{OK. Let's leave Definition~\ref{def:ranking function} as it is.}
\fi
We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.

@@

\begin{remark}
Lemma~\ref{lem:rank-PAST} is a slight extension of \cite[Lem.~5.5]{DBLP:conf/popl/FioritiH15},
which asserts the same result but for the specific stopping time $T : \omega \mapsto \min\set{n \mid Y_n(\omega) = 0}$.
\end{remark}

@@

%\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
The ranking supermartingale condition is satisfied essentially by a case analysis on the type of redex (Lemma~\ref{lem:key rankable}).

@@

\begin{proof}
Let $f$ be a ranking function on $M$.
For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (Lemma~\ref{lem:TnBounded}),
and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (Theorem~\ref{thm:rankable and strict rankable}),
it follows from the Optional Sampling Theorem~\ref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$.
Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (Lemma~\ref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
Therefore, by Lemma~\ref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
%For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
%It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
Statement (ii) follows at once from Theorem~\ref{thm:rankable and strict rankable} and Lemma~\ref{lem:rank-PAST}.
\end{proof}

Thus the method of (strict) ranking function is sound for proving (positive) a.s.~termination of PPCF programs.
It is in fact also complete in a sense: if $\expect{T^\tY_N} < \infty$ for all $N \in \mathit{Rch}(M)$ then $M$ is rankable (Theorem~\ref{thm:minimal}).

### 2021-10-01 probabilisticProgrammingMartingales.tex
" \ref" -> "~\ref"

\begin{theorem} \label{thm:minimal}
Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
\end{theorem}
\begin{proof}
Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$. The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$. The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
%Theorem \ref{rankable implies ast}) 
\Cref{thm:rankable and strict rankable});
therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost surely finite.
\end{proof}

@@

\begin{theorem}  \label{thm:antitone partial implies rankable}
  Every antitone partial ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

@@

\paragraph{Completeness}
It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this. It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem \ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).

@@

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. \lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma \ref{lem:independentSamples} for example.}
\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\)

@@

Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma \ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$. Let the position of the reduction $M \to N$ be $\beta$. 
\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions. 
The reductions derived from $M \to N$ can be identified as follows:

@@

\begin{lemma} \label{lem:redexDestroyed}
If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
\end{lemma}
\begin{proof}
Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma \ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$. $M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
\end{proof}

@@

In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$. If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample. For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent. The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$. Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$. In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.

@@

\begin{theorem} \label{thm:CbvIsTerminatingest}
If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
\end{theorem}
\begin{proof}
  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$. Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value. By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order. For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction). Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$. To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions. Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related. If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially. If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches. Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$. The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.

@@

All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too. The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies. They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy). For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x. O]$, and $r(N)$ being undefined for $N$ being a value.

>>>>

\begin{theorem} \label{thm:minimal}
Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a value}]$, if it exists, is the least of all possible ranking functions of $M$.
\end{theorem}
\begin{proof}
Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$. The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$. The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
%Theorem~\ref{rankable implies ast}) 
\Cref{thm:rankable and strict rankable});
therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem~\ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost surely finite.
\end{proof}

@@

\begin{theorem}  \label{thm:antitone partial implies rankable}
  Every antitone partial ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem~\ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

@@

\paragraph{Completeness}
It seems very likely that this method is complete in the case that every reachable term is AST, but we have been unable to actually prove this. It is certainly at least capable of proving the termination of terms which terminate arbitrarily slowly (in the sense that there is no similar limitation to Theorem~\ref{thm:rankable implies termination}, which can only prove termination of $\tY$-PAST terms).

@@

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $Sk(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. \lo{Do we actually use the notation $S[r]$ in what follows?} \akr{Yes, in Lemma~\ref{lem:independentSamples} for example.}
\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\)

@@

Let $X$ be the call-by-value reduction sequence related to both $N'_1$ and $N'_2$ by $\sim_{c,M}$ given by Lemma~\ref{canonicalCousins}. 
As $M \to N$ is the first reduction in the sequence $N'_1$, it is the last to be affected by the $\sim_{c,M}$ sequence $N'_1 \sim_{c,M}^* X$ given by Lemma~\ref{canonicalCousins} therefore it can be split into $N'_1 \sim_{c,M}^* Y_1 \sim_{c,M}^* X$ where $Y_1$ is in CBV order except possibly for its first reduction, which is still $M \to N$. Let the position of the reduction $M \to N$ be $\beta$. 
\akr{TODO: fill in the details that are missing here. It shouldn't be too hard, it's just a bit sketchy at present.} 
The rearrangement of the reduction sequence $Y_1 \sim_{c,M} X$ consists of moving the reduction at $\beta$ down past the other reductions in $Y_1$, possibly duplicating or deleting it in the process, but not affecting the positions or the correct order for any of the other reductions. 
The reductions derived from $M \to N$ can be identified as follows:

@@

\begin{lemma} \label{lem:redexDestroyed}
If $M \to N$, with the redex at position $\alpha$, then no position in any term reachable from $N$ is related by $\sim^*$ to $(M,\alpha)$.
\end{lemma}
\begin{proof}
Suppose on the contrary that $(M,\alpha) \sim^* (A,\beta)$, where $N \to^* A$, then by Lemma~\ref{simpDoesn'tMix}, $(M,\alpha) \sim_p^* (A,\beta)$. $M \neq A$ therefore $(M,\alpha) \sim_p$ some position in $N$, but in all the cases of the definition of $\sim_p$, no position in the child term is related to the position of the redex.
\end{proof}

@@

In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$. If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample. For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent. The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$. Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma~\ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$. In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.

@@

\begin{theorem} \label{thm:CbvIsTerminatingest}
If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
\end{theorem}
\begin{proof}
  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$. Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value. By Lemma~\ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order. For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction). Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$. To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions. Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma~\ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related. If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially. If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches. Similarly to Lemma~\ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$. The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.

@@

All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too. The proofs of Theorems~\ref{thm:rankable implies termination},~\ref{thm:partial implies rankable},~\ref{thm:antitone rankable implies termination} and~\ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies. They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem~\ref{thm:de groote} applies equally to any reduction strategy). For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x. O]$, and $r(N)$ being undefined for $N$ being a value.
